Appendix to "Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model

### 2023-06-19 src/docs_paper/appendix.Rmd
Reworking appendix, some main text

---
title: Appendix to "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model"
author:
  - Adam Howes^[Department of Mathematics, Imperial College London]
  - Alex Stringer^[Department of Statistics and Actuarial Science, University of Waterloo]
  - Seth R. Flaxman^[Department of Computer Science, Oxford University]
  - Jeffrey W. Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
output:
  pdf_document:
    toc: yes
    number_sections: yes
    keep_tex: yes
    includes:
      in_header: preamble.tex
bibliography: citations.bib
cls: imsart.cls
---

@@

```{=tex}
\setcounter{table}{0}  
\renewcommand{\thetable}{S\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{section}{0} 
\renewcommand{\thesection}{S\arabic{section}}
```

@@

In this section we describe the simplified Naomi model [@eaton2021naomi] considered in the main text in more complete detail.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
We let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.
In the main text we use a more concise, but less descriptive, approach.

@@

We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with length $\dim(u)$.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].

@@

```{=tex}
\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXX}
\toprule
& Model component & Latent field & Hyperparameter \\ 
\midrule
\ref{sec:hiv-prev} & HIV prevalence & $22 + 5n$ & 9 \\ 
\ref{sec:art-cov} & ART coverage & $25 + 5n$ & 9 \\ 
\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
\ref{sec:anc-test} & ANC testing & $2 + 2n$ & 2 \\ 
\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
& Total & $51 + 14n$ & 24 \\ 
\bottomrule
\end{tabularx}
\label{tab:latent-hyper-size}
\caption{The numer of latent field parameters and hyperparameters in each model section.}
\end{table}
```

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}. 
$$
The term $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept, and $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women.
The terms $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, and $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$.
The terms $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, and $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$.
The terms $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence.
Finally, $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
The two BYM2 random effects are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components
\begin{align*}
u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
\end{align*}
We use half-normal priors for the standard deviation terms
$$
\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
$$
uniform priors for the AR1 correlation parameters
$$
\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
$$
and beta priors for the BYM2 proportion parameters
$$
\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
$$

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
$$
where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults and $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women.
The term
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}
$$
is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}
$$
is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation].
The terms $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects.
Finally, $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output^[Note that these outputs are the only source of age structure for this part of the model. Furthermore, Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80. A consequence is that the posterior over new infections in these age groups is exactly zero, by definition. We remove these identically zero posteriors from any inference comparisons.]

@@

In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
Here we fix $\beta^\psi = u_x^\psi = 0$ such that $\Psi_{x, a}$ are simply constants.

@@

For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
$$
y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
$$
To model these observations, we use following three binomial working likelihoods
\begin{align*}
y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align*}
The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
$$
\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
$$
where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.

@@

@@

```{=tex}

@@

```

@@

```{r rhat, fig.cap="The potential scale reduction factor compared between- and within- estimates of univariate parameters. It is recommended against to use the results if the value is less than 1.05."}
knitr::include_graphics("depends/rhat.png")
```

```{r ratio, fig.cap="The ESS ratio measures how efficient sampling is."}
knitr::include_graphics("depends/ratio.png")
```

```{r ess, fig.cap="The effective number of samples varied substantially."}
knitr::include_graphics("depends/ess.png")
```

@@

# AGHQ and PCA-AGHQ

@@

```{r scree, fig.cap="Most variation could be explained by fewer than the full 24 dimensions."}
knitr::include_graphics("depends/tv-plot.png")
```

```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank (24) matrix."}
knitr::include_graphics("depends/reduced-rank-plot.png")
```

@@

## KS tests

@@

@@

# Laplace marginals algorithm

In earlier work, we used a full Laplace approximation to the latent field marginals.
The algorithm for obtaining this approximation is detailed here.

>>>>

---
title: Appendix to "Fast approximate Bayesian inference of HIV indicators using the Naomi small-area estimation model"
author:
  - Adam Howes^[Department of Mathematics, Imperial College London]
  - Alex Stringer^[Department of Statistics and Actuarial Science, University of Waterloo]
  - Seth R. Flaxman^[Department of Computer Science, Oxford University]
  - Jeffrey W. Eaton^[Harvard T.H. Chan School of Public Health, Harvard University]
output:
  pdf_document:
    toc: yes
    number_sections: yes
    keep_tex: yes
    includes:
      in_header: preamble.tex
bibliography: citations.bib
cls: imsart.cls
---

@@

\setcounter{table}{0}  
\renewcommand{\thetable}{S\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{section}{0} 
\renewcommand{\thesection}{S\arabic{section}}

@@

Here, we describe the simplified Naomi model [@eaton2021naomi] in more detail.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
We then use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
Let $\sum_{\{x\}}$ be shorthand for $\sum_{x \in \{x\}}$, analogously for $s$ and $a$.

@@

We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with length $\dim(u)$.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean, calculated using the `survey` \textsc{R} package [@lumley2004analysis], is given by
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.

@@

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXX}
\toprule
& Model component & Latent field & Hyperparameter \\ 
\midrule
\ref{sec:hiv-prev} & HIV prevalence & $22 + 5n$ & 9 \\ 
\ref{sec:art-cov} & ART coverage & $25 + 5n$ & 9 \\ 
\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
\ref{sec:anc-test} & ANC testing & $2 + 2n$ & 2 \\ 
\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
& Total & $51 + 14n$ & 24 \\ 
\bottomrule
\end{tabularx}
\caption{The numer of latent field parameters and hyperparameters in each model section.}
\label{tab:latent-hyper-size}
\end{table}

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
\begin{equation}
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}. \label{eq:prev}
\end{equation}
Table \ref{tab:prev} provides a description of the terms included in Equation \ref{eq:prev}.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\rho_0$ & $\mathcal{N}(0, 5)$ & Intercept \\
$\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
$u^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
$u_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
$u^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
$u_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
$u_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
$\eta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV prevalence (Equation \ref{eq:prev}).}
\label{tab:prev}
\end{table}

The two BYM2 random effects $u^\rho_x$ and $u_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components, respectively
\begin{align*}
u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
\end{align*}
We use half-normal priors for the standard deviation terms
$$
\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
$$
uniform priors for the AR1 correlation parameters
$$
\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
$$
and beta priors for the BYM2 proportion parameters
$$
\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
$$

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
\begin{equation}
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda. \label{eq:inc}
\end{equation}
Table \ref{tab:inc} provides a description of the terms included in Equation \ref{eq:inc}.

<!-- \citet{eaton2019estimation} -->

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\lambda_0$ & $\mathcal{N}(0, 5)$ & Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults \\
$\beta_S^{\lambda, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The log incidence rate ratio for men compared to women \\
$\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}$ & $-$ & The HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences \\
$\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}$ & $-$ & The ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages \\
$\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
$u_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
$\eta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}). Note that the only source age structure for this part of the model are $\eta^\lambda_{R_x, s, a}$. As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition. We remove these identically zero posteriors from any later inference comparisons.}
\label{tab:inc}
\end{table}

@@

In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi,
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
Here we fix $\beta^\psi = u_x^\psi = 0$ such that $\Psi_{x, a}$ are simply constants.

@@

For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the most recent household survey furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\theta}_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}$, and observed number of cases
$$
y^{\theta}_{\{x\}, \{s\}, \{a\}} = m^{\theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}} \in \mathbb{R}.
$$
To model these observations, we use following three binomial working likelihoods
\begin{align*}
y^{\rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
y^{\alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}, \\
y^{\kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align*}
The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
$$
\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
$$
where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.

@@

<!-- Create DAG representing Naomi model. This would likely go into the appendix, so use that notation -->

@@

@@

@@

```{r rhat, fig.cap="The potential scale reduction factor compares between- and within- estimates of univariate parameters. It is recommended only to use NUTS results if the value is less than 1.05, which it is for all parameters."}
knitr::include_graphics("depends/rhat.png")
```

```{r ratio, fig.cap="The efficiently, as measured by the ESS ratio, of the NUTS sampler was poor."}
knitr::include_graphics("depends/ratio.png")
```

```{r ess, fig.cap="The effective number of samples we obtained varied substantially between parameters."}
knitr::include_graphics("depends/ess.png")
```

@@

# AGHQ and PCA-AGHQ details

## Additional figures

@@

```{r scree, fig.cap="Most variation could be explained by far fewer than the full 24 dimensions."}
knitr::include_graphics("depends/tv-plot.png")
```

```{r reduced-rank, fig.cap="The reduced rank (8) matrix approximation to the Hessian is visually similar to the full rank matrix."}
knitr::include_graphics("depends/reduced-rank-plot.png")
```

## Estimated normalising constant comparison

@@

## Point estimates

## Distribution tests

@@

## Pareto-smoothed importance sampling

## Maximum mean discrepancy

@@

# Algorithm using Laplace latent field marginals
